---
title: "final_project_epi204"
output: html_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing

```{r}
#load library
library(geepack)
library(lmtest)
library(tidyverse)
library(readr)
library(tableone)

```

```{r}
nhanes2 <- read_csv("NHANES2.csv")
```



## Model 1: Cox PH Regression

```{r}
# Example Cox PH code
#fit.cox.1<-coxph(Surv(AGEDEATH, DEATH) ~ AGEYRS + COVARIATES, data = nhanes2, ties='efron')
#summary(fit.cox.1)

#checking Schoenfeld Residuals
#check.pha<-cox.zph(fit.cox.1)
#ggcoxzph(check.pha[1])

```


## Model 2: Poisson Model


```{r}
#Example Poisson code
#fit.pois1 <- glm(DEATH ~ AGEYRS + COVARIATES, data=nhanes2, offset=OFFSET, family=poisson(link='log')) #specify COVARIATES and offset to run this model
#summary(fit.pois1)
```



## Table 1 
Describing relevant basic characteristics of study participants according to exposure categories of your choice

```{r}
nhanes3<-nhanes2 %>% 
  mutate(sex=factor(SEX, labels=c("male", "female"), levels=c(1,2)), #relabeling sex
         niacin_cat=ifelse(NIACIN>=16 & SEX=="male", "Adequate", "Inadequate")) 
      #creating Niacin categories based on DV: The DV for niacin is 16 mg for adults and children age 4        years and older

vars <- (c("sex")) # include more covariates as needed
t1 <- CreateTableOne(vars = vars, data = nhanes3, strata = "niacin_cat")
options(digits = 2)
print(t1, showAllLevels = TRUE)
```


## Table 2
Present the age-adjusted and multivariable-adjusted results of the association generated from both of your models

```{r}


```



## Table 3
Give the multivariable-adjusted results in each category of the effect modifier (sex), and p-values for interaction (using the primary model of your choice only)

```{r}

```







