---
title: "final_project_epi204"
output: html_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing

```{r setup, include = FALSE}
# # Clear environment 
rm(list = ls())

# Load dependencies
library(geepack)
library(lmtest)
library(tidyverse)
library(readr)
library(tableone)
library(dplyr)
```

```{r}
DirPath <- getwd()
nhanes <- read_csv(paste0(DirPath, "/data/NHANES2.csv"))
```

```{r}
# Data setup

## Data checks
### Confirm one row per person
length(unique(nhanes$SEQNO)) == nrow(nhanes)
```

```{r}
## Select variables of interest

nhanes_analytic <- nhanes |>
  select(
    # admin vars
    c(SEQNO, AGEYRS, AGEDIE, 
      LAST_YR, LAST_MO, LAST_DAY, 
      BORN_YR, BORN_MO, 
      DEATH, DIE_YR, DIE_MO,
    # exposure
      NIACIN,
    # chronic conditions/comorbidities
      ANGINA, ARTHRITI, ASPIRIN , EVPREG, HTN_REP, STROKE, 
    # sociodempgraphic variables
      MARRY, RACE, SEX, URBAN, SCHOOL,
    # lifestyle variables
      BOOZE, RECEX, AVGSMK, HEIGHT, WT)
  )

### Review covariate missingness
library(naniar)
gg_miss_var(nhanes_analytic)
miss_var_summary(nhanes_analytic)

# Checking something funky going on with time variables
# table(!is.na(nhanes_analytic$BORN_YR))
# table(!is.na(nhanes_analytic$LAST_YR))

# check <- nhanes_analytic |>
#   filter(is.na(BORN_YR),
#          is.na(LAST_YR))

# AB: AGEDIE, DIE_YR are not well populated (~80% missing) though DEATH outcome is not that bad (~17.8% missing); just checked and it's because AGEDIE and DIE_YR are listed NA for those who have DEATH = 1; let's keep even with missing - these are not covars

table(nhanes$AGEDIE, nhanes$DEATH, useNA = "ifany") # AB: no data for those 2,008 we're excluding. those with NA for AGEDIE are coded as 0 for DEATH

table(nhanes_analytic$EVPREG, nhanes_analytic$SEX)
# EVPREG missing only for assigned male at birth - update variable below to counter missingnnes

# AB: Test out N for final analytical sample if we keep only participants with covariates of interest (excepting those highlighted above)
nhanes_final <- nhanes_analytic |>
  mutate(EVPREG = ifelse(is.na(EVPREG) & SEX == 1, "Not Applicable",
                         EVPREG)) |>
  filter(!is.na(NIACIN),
         !is.na(ANGINA),
         !is.na(STROKE),
         !is.na(HTN_REP),
         !is.na(LAST_YR),
         !is.na(BORN_YR),
         !is.na(DEATH),
         !is.na(URBAN),
         !is.na(SCHOOL),
         !is.na(ARTHRITI),
         !is.na(EVPREG)) 

nrow(nhanes_final) #8,459 patients
```

```{r}
# Summarize niacin
summary(nhanes_final$NIACIN, useNA = "ifany")

# Create niacin quartiles
niacin_qval <- nhanes_final %>%
  summarise(p25 = quantile(NIACIN, probs = 0.25),
            p50 = quantile(NIACIN, probs = 0.50),
            p75 = quantile(NIACIN, probs = 0.75))

niacin_qval_vec <- as.vector(niacin_qval)

nhanes_final <- nhanes_final |>
  mutate(niacin_q = ifelse(
    nhanes_final$NIACIN < niacin_qval$p25, 1, 
                         ifelse(
                           nhanes_final$NIACIN >= niacin_qval$p25 & nhanes_final$NIACIN < niacin_qval$p50, 2, 
                                ifelse(
                                  nhanes_final$NIACIN >= niacin_qval$p50 & nhanes_final$NIACIN < niacin_qval$p75, 3, 
                                       ifelse(# AB: this part of the code was leaving out two participants exactly at the cutoff of p75; changed > to >= 
                                         nhanes_final$NIACIN >= niacin_qval$p75, 4, NA)))))

# AB: Check with cut
# qs <- quantile(nhanes_final$NIACIN,
#                probs = seq(0, 1, 0.25),
#                na.rm = TRUE)
# 
# # test <- nhanes_final
# nhanes_final$niacin_q_test <- cut(nhanes_final$NIACIN,
#             breaks = qs,
#             include.lowest = FALSE,
#             labels = c(1, 2, 3, 4))
# 
# table(nhanes_final$niacin_q_test, useNA = "ifany")
# table(nhanes_final$niacin_q, useNA = "ifany") 
# 
# subset_unequal <- nhanes_final |>
#   filter(niacin_q != niacin_q_test) |>
#   select(c(SEQNO, NIACIN, niacin_q, niacin_q_test))
# 
# nhanes_final <- nhanes_final |>
#   select(-c(niacin_q_test))
```

```{r}
# Variable Operationalization
# BMI, person-time for Poisson analysis
nhanes_final_2 <- nhanes_final |>
mutate(
  # using age as the time scale; person-time accumulates in age years
  # time zero is age at entry
  age_tstart = AGEYRS,
  # end of follow up is censoring or death
  # age_tend = LAST_YR - BORN_YR,
  age_tend = if_else(!is.na(AGEDIE), AGEDIE, LAST_YR),
  person_time = age_tend - age_tstart,
  event = if_else(!is.na(AGEDIE) & DEATH == 1, 1, 0),
  bmi = WT/(HEIGHT/100)^2
)

nhanes_model <- nhanes_final_2 |>
  filter(
    is.na(person_time),
    person_time > 0
  )

summary(nhanes_model$person_time) # max 63 years of FU
```


## Model 1: Poisson Model (Primary)


```{r}
#Example Poisson code
#fit.pois1 <- glm(DEATH ~ AGEYRS + COVARIATES, data=nhanes, offset=OFFSET, family=poisson(link='log')) #specify COVARIATES and offset to run this model
#summary(fit.pois1)
```

## Model 2: Cox PH Model (Sensitivity)

```{r}
# Example Cox PH code
#fit.cox.1<-coxph(Surv(AGEDEATH, DEATH) ~ AGEYRS + COVARIATES, data = nhanes, ties='efron')
coxph(
  Surv(entry_age, exit_age, event) ~ exposure + covariates,
  data = df
)
#summary(fit.cox.1)

#checking Schoenfeld Residuals
#check.pha<-cox.zph(fit.cox.1)
#ggcoxzph(check.pha[1])

```


## Table 1 
Describing relevant basic characteristics of study participants according to exposure categories of your choice

```{r}
nhanes2<-nhanes %>% 
  mutate(sex=factor(SEX, labels=c("male", "female"), levels=c(1,2)), 
         age= AGEYRS,
         angina=factor(ANGINA, levels=c(0,1)),
         arthritis = factor(ARTHRITI, levels=c(0,1)),
         aspirin = factor(ASPIRIN, levels=c(0,1)),
         ever_pregnant=factor(EVPREG,levels=c(0,1)), # might need to show it only among females
         hypertension=factor(HTN_REP,levels=c(0,1)),
         marital_status=factor(MARRY, 
                               labels=c("Under 17", "Married", "Widowed",
                                        "Divorced", "Separated", "Never Married",
                                        "Blank, but applicable"), 
                               levels=c(1,2,3,4,5,6,8)),
         race=factor(RACE,
                     labels=c("White","Black","Other"),
                     levels=c(1,2,3)),
         urban=factor(URBAN,
                      labels=c("Yes", "No"), levels=c(1,2)),
         school=factor(SCHOOL,
                       labels=c("Elementary", "Some high school", 
                                      "High school", "Some college", "College"), 
                               levels=c(1,2,3,4,5)),
         alcohol=as.numeric(BOOZE),
         recex=factor(RECEX, labels=c("High","Medium","Low"),levels=c(1,2,3)),
         smoking=as.numeric(AVGSMK))

vars <- (c("sex", "age","angina","arthritis", 
           "aspirin","ever_pregnant","hypertension",
           "marital_status", "race", "urban",
           "school", "alcohol", 
           "recex", "smoking",
           "bmi")) # include more covariates as needed

# Table 1 needs complete case

t1 <- CreateTableOne(vars = vars, data = nhanes2, strata = as.factor("niacin_q"))
options(digits = 2)
print(t1, showAllLevels = TRUE)
```


## Table 2
Present the age-adjusted and multivariable-adjusted results of the association generated from both of your models

```{r}


```



## Table 3
Give the multivariable-adjusted results in each category of the effect modifier (sex), and p-values for interaction (using the primary model of your choice only)

```{r}

```







