---
title: "final_project_epi204"
output: html_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing

```{r setup, include = FALSE}
# # Clear environment 
rm(list = ls())

# Load dependencies
library(geepack)
library(lmtest)
library(tidyverse)
library(readr)
library(tableone)
library(dplyr)
```

```{r}
DirPath <- getwd()
nhanes <- read_csv(paste0(DirPath, "/data/NHANES2.csv"))
```

```{r}
## Data setup
## Data Structure
### confirming one row per person
length(unique(nhanes$SEQNO)) == nrow(nhanes)

# Selecting variables of interest:

nhanes <- nhanes |>
  select(
    # admin vars
    c(SEQNO, AGEYRS, AGEDIE, LAST_YR, BORN_YR, DEATH, DIE_YR,
    # exposure
      NIACIN,
    # chronic conditions/comorbidities
      ANGINA, ARTHRITI, ASPIRIN , EVPREG, HTN_REP, STROKE, 
    # sociodempgraphic variables
      MARRY, RACE, SEX, URBAN, SCHOOL,
    # lifestyle variables
      BOOZE, RECEX, AVGSMK, HEIGHT, WT)
  )

### Missingness

# Niacin Quantiles

niacin_qval <- nhanes %>%
  filter(!is.na(NIACIN)) %>%
  summarise(p25 = quantile(NIACIN, probs = 0.25),
            p50 = quantile(NIACIN, probs = 0.50),
            p75 = quantile(NIACIN, probs = 0.75))

nhanes$niacin_q <- ifelse(nhanes$NIACIN < niacin_qval$p25, 1, 
                         ifelse(nhanes$NIACIN >= niacin_qval$p25 & nhanes$NIACIN < niacin_qval$p50, 2, 
                                ifelse(nhanes$NIACIN >= niacin_qval$p50 & nhanes$NIACIN < niacin_qval$p75, 3, 
                                       ifelse(nhanes$NIACIN > niacin_qval$p75, 4, NA))))
table(nhanes$niacin_q)

# Variable Operationalization
## BMI, person-time for Poisson analysis
nhanes <- nhanes |>
  mutate(
    # using age as the time scale
    # time zero is age at entry
    age_tstart = AGEYRS,
    # end of follow up is censoring or death
    age_tend = LAST_YR - BORN_YR,
    pt = age_tend - age_tstart,
    bmi = WT/(HEIGHT/100)^2
  )

# Checking something funky going on with time variables
table(!is.na(nhanes$BORN_YR))
table(!is.na(nhanes$LAST_YR))

nhanes_check <- nhanes |>
  filter(is.na(BORN_YR),
         is.na(LAST_YR))

nhanes_model <- nhanes |>
  filter(
    is.na(pt),
    pt > 0
  )

#a quick look at covariates with missingness
library(naniar)
gg_miss_var(nhanes)

```




## Model 1: Cox PH Regression

```{r}
# Example Cox PH code
#fit.cox.1<-coxph(Surv(AGEDEATH, DEATH) ~ AGEYRS + COVARIATES, data = nhanes, ties='efron')
#summary(fit.cox.1)

#checking Schoenfeld Residuals
#check.pha<-cox.zph(fit.cox.1)
#ggcoxzph(check.pha[1])

```


## Model 2: Poisson Model


```{r}
#Example Poisson code
#fit.pois1 <- glm(DEATH ~ AGEYRS + COVARIATES, data=nhanes, offset=OFFSET, family=poisson(link='log')) #specify COVARIATES and offset to run this model
#summary(fit.pois1)
```



## Table 1 
Describing relevant basic characteristics of study participants according to exposure categories of your choice

```{r}
nhanes2<-nhanes %>% 
  mutate(sex=factor(SEX, labels=c("male", "female"), levels=c(1,2)), 
         age= AGEYRS,
         angina=factor(ANGINA, levels=c(0,1)),
         arthritis = factor(ARTHRITI, levels=c(0,1)),
         aspirin = factor(ASPIRIN, levels=c(0,1)),
         ever_pregnant=factor(EVPREG,levels=c(0,1)), # might need to show it only among females
         hypertension=factor(HTN_REP,levels=c(0,1)),
         marital_status=factor(MARRY, 
                               labels=c("Under 17", "Married", "Widowed",
                                        "Divorced", "Separated", "Never Married",
                                        "Blank, but applicable"), 
                               levels=c(1,2,3,4,5,6,8)),
         race=factor(RACE,
                     labels=c("White","Black","Other"),
                     levels=c(1,2,3)),
         urban=factor(URBAN,
                      labels=c("Yes", "No"), levels=c(1,2)),
         school=factor(SCHOOL,
                       labels=c("Elementary", "Some high school", 
                                      "High school", "Some college", "College"), 
                               levels=c(1,2,3,4,5)),
         alcohol=as.numeric(BOOZE),
         recex=factor(RECEX, labels=c("High","Medium","Low"),levels=c(1,2,3)),
         smoking=as.numeric(AVGSMK))

vars <- (c("sex", "age","angina","arthritis", 
           "aspirin","ever_pregnant","hypertension",
           "marital_status", "race", "urban",
           "school", "alcohol", 
           "recex", "smoking",
           "bmi")) # include more covariates as needed

# Table 1 needs complete case

t1 <- CreateTableOne(vars = vars, data = nhanes2, strata = as.factor("niacin_q"))
options(digits = 2)
print(t1, showAllLevels = TRUE)
```


## Table 2
Present the age-adjusted and multivariable-adjusted results of the association generated from both of your models

```{r}


```



## Table 3
Give the multivariable-adjusted results in each category of the effect modifier (sex), and p-values for interaction (using the primary model of your choice only)

```{r}

```







