---
title: "EPI 204: Final Project"
author: "Andra Boca, Parker Tope, Yvonne Chen, Billy Huang, Camryn Dixon"
output: html_document
date: "2026-04-30"
---

## Data Processing

```{r setup, include = FALSE}
# # Clear environment 
rm(list = ls())

# Load dependencies
library(geepack)
library(lmtest)
library(tidyverse)
library(readr)
library(tableone)
library(dplyr)
library(survival)
library(survminer)
library(splines)
knitr::opts_chunk$set(echo = TRUE)
library(lmtest)
```

```{r, results = FALSE}
DirPath <- getwd()
nhanes <- read_csv(paste0(DirPath, "/data/NHANES2.csv"))
### Confirm one row per person
length(unique(nhanes$SEQNO)) == nrow(nhanes)
```

## Selecting variables of interest

```{r}
## Select variables of interest
nhanes_analytic <- nhanes |>
  select(
    # admin vars
    c(SEQNO, AGEYRS, AGEDIE, 
      LAST_YR, LAST_MO, LAST_DAY, 
      BORN_YR, BORN_MO, 
      DEATH, DIE_YR, DIE_MO,
    # exposure
      NIACIN,
    # chronic conditions/comorbidities
      ANGINA, ARTHRITI, ASPIRIN , EVPREG, HTN_REP, STROKE, 
    # sociodempgraphic variables
      MARRY, RACE, SEX, URBAN, SCHOOL,
    # lifestyle variables
      BOOZE, RECEX, AVGSMK, HEIGHT, WT)
  )
```

## Missingness of variables
```{r}
### Review covariate missingness
library(naniar)
gg_miss_var(nhanes_analytic)

# AB: AGEDIE, DIE_YR are not well populated (~80% missing) though DEATH outcome is not that bad (~17.8% missing); just checked and it's because AGEDIE and DIE_YR are listed NA for those who have DEATH = 1; let's keep even with missing - these are not covars

nhanes_analytic_missingness <- nhanes_analytic |>
  select(
    NIACIN,
    ANGINA,
    STROKE,
    HTN_REP,
    LAST_YR,
    BORN_YR,
    DEATH,
    URBAN,
    SCHOOL,
    ARTHRITI
    )

# Exporting table of missingness proportions for vars of interest
miss_var_summary(nhanes_analytic)
nhanes_analytic_missingness_table <- miss_var_summary(nhanes_analytic_missingness) |>
  as.data.frame() |>
  mutate(
    n_miss = as.numeric(n_miss),
    pct_miss = as.numeric(pct_miss)
  )

write.csv(nhanes_analytic_missingness_table, "nhanes_missing.csv", row.names = FALSE)

table(nhanes$AGEDIE, nhanes$DEATH, useNA = "ifany") # AB: no data for those 2,008 we're excluding. those with NA for AGEDIE are coded as 0 for DEATH

table(nhanes_analytic$EVPREG, nhanes_analytic$SEX)
# EVPREG missing only for assigned male at birth; keep but may only be usable for analysis among those assigned female at birth? Check notes below.

# AB: Test out N for final analytical sample if we keep only participants with covariates of interest (except those highlighted above)
nhanes_final <- nhanes_analytic |>
  filter(!is.na(NIACIN),
         !is.na(ANGINA),
         !is.na(STROKE),
         !is.na(HTN_REP),
         !is.na(LAST_YR),
         !is.na(BORN_YR),
         !is.na(DEATH),
         !is.na(URBAN),
         !is.na(SCHOOL),
         !is.na(ARTHRITI)) 

nrow(nhanes_final) #8,490 patients
```

## Niacin Quartile Operationalization

```{r}
# Summarize niacin
summary(nhanes_final$NIACIN, useNA = "ifany")

# Plot distribution exposure
ggplot(nhanes_final,
       aes(x = NIACIN)) + 
  geom_histogram()

# Create niacin quartiles
niacin_qval <- nhanes_final %>%
  summarise(p25 = quantile(NIACIN, probs = 0.25),
            p50 = quantile(NIACIN, probs = 0.50),
            p75 = quantile(NIACIN, probs = 0.75))

niacin_qval_vec <- as.vector(niacin_qval)

nhanes_final <- nhanes_final |>
  mutate(niacin_q = ifelse(
    nhanes_final$NIACIN < niacin_qval$p25, 1, 
                         ifelse(
                           nhanes_final$NIACIN >= niacin_qval$p25 & nhanes_final$NIACIN < niacin_qval$p50, 2, 
                                ifelse(
                                  nhanes_final$NIACIN >= niacin_qval$p50 & nhanes_final$NIACIN < niacin_qval$p75, 3, 
                                       ifelse(# AB: this part of the code was leaving out two participants exactly at the cutoff of p75; changed > to >= 
                                         nhanes_final$NIACIN >= niacin_qval$p75, 4, NA)))))

nhanes_final$niacin_q <- as.factor(nhanes_final$niacin_q)
```

## BMI, person-time, sex

```{r}
# Variable Operationalization
# BMI, person-time
nhanes_final_2 <- nhanes_final |>
mutate(
  # PRT: using age as the time scale; person-time accumulates in age years
  # PRT: time zero is age at entry
  age_tstart = AGEYRS,
  # PRT: end of follow up is censoring or death
  # age_tend = LAST_YR - BORN_YR,
  age_tend = if_else(!is.na(AGEDIE), AGEDIE, LAST_YR - BORN_YR),
  person_time = age_tend - age_tstart,
  event = if_else(!is.na(AGEDIE) & DEATH == 1, 1, 0),
  BMI = WT/(HEIGHT/100)^2,
  # 1 = female, 0 = male
  SEX = if_else(SEX == 2, 1, 0)
)

nhanes_model <- nhanes_final_2 |>
  filter(
    !is.na(person_time),
    person_time > 0
  )

summary(nhanes_model$person_time) # PRT: min 1 and max 17 years of FU

```

## Class of variables

```{r}
nhanes_final$niacin_q <- as.factor(nhanes_final$niacin_q)
nhanes_final$RACE <- as.factor(nhanes_final$RACE)
nhanes_final$SEX <- as.factor(nhanes_final$SEX)
nhanes_final$SCHOOL <- as.factor(nhanes_final$SCHOOL)
nhanes_final$MARRY <- as.factor(nhanes_final$MARRY)
nhanes_final$URBAN <- as.factor(nhanes_final$URBAN)
nhanes_final$ANGINA <- as.factor(nhanes_final$ANGINA)
nhanes_final$ARTHRITI <- as.factor(nhanes_final$ARTHRITI)
nhanes_final$ASPIRIN <- as.factor(nhanes_final$ASPIRIN)
nhanes_final$HTN_REP <- as.factor(nhanes_final$HTN_REP)
nhanes_final$STROKE <- as.factor(nhanes_final$STROKE)
```

```{r}
# Covariates list (so it's easy to update here only once if changed and automatically update all models downstream)
cov_basic <- c(
  "SEX",
  "RACE",
  "SCHOOL"
)

cov_full <- c(
  "SEX",
  "RACE",
  "SCHOOL",
  "MARRY",
  "URBAN",
  "BOOZE",
  "AVGSMK",
  "BMI",
  "ANGINA",
  "ARTHRITI",
  "ASPIRIN",
  "HTN_REP",
  "STROKE"
)
```

## Primary Analysis: Poisson Model

### Generating one row per follow-up person-year dataset for Poisson
```{r, results = FALSE}
nhanes_final_poisson <- nhanes_model |>
  uncount(person_time, .id = "pt_year", .remove = FALSE) |>
  mutate(
    # generating individuals' age in year of follow up
    age_in_year = AGEYRS + pt_year - 1,
    # generating indicator var for death for one row per year of FU data
    death_in_year = ifelse((event == 1 & pt_year == person_time), 1, 0),
    log_pt = log(1)
)

# soundness check
n_distinct(nhanes_final_poisson$SEQNO)
# 8461 individuals

# SOUNDNESS CHECK
nhanes_final_poisson |>
  filter(DEATH == 1) |>
  summarise(
    deaths_pre_split = n_distinct(SEQNO),
    deaths_post_split = sum(death_in_year),
    max_fu = max(age_tend - age_in_year)
  )
```

### Model 1.1 Age-adjusted model

```{r}
fit.pois1 <- glm(death_in_year ~ # exposure 
                   niacin_q + 
                   # modelling age in interval
                   ns(age_in_year, df = 4),
                 data=nhanes_final_poisson, 
                 offset=log_pt, 
                 family=poisson(link='log'))
summary(fit.pois1)
```

```{r}
## Poisson Model 1.1 Robust SEs
# Get robust variance-covariance matrix
robust_vcov1 <- sandwich::vcovHC(fit.pois1, type = "HC3")
tidy_model1 <- broom::tidy(fit.pois1, conf.int = TRUE, conf.level = 0.95)

# Replace standard errors with robust ones
robust_results1 <- coeftest(fit.pois1, vcov = robust_vcov1)
tidy_model1$std.error <- robust_results1[, "Std. Error"]
# Calculate RR (95% CI) with robust standard errors
tidy_poisson1 <- tidy_model1 %>%
  mutate(
    RR = exp(estimate),
    RR_low = exp(estimate - 1.96 * std.error),
    RR_high = exp(estimate + 1.96 * std.error)
  )
tidy_poisson1
```

### Model 1.2: Minimally adjusted model
```{r}
form.pois2 <- as.formula(
  paste(
    "death_in_year ~ niacin_q +",
    paste(cov_basic, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois2 <- glm(form.pois2,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))
summary(fit.pois2)
```

```{r}
## Poisson Model 1.2 Robust SEs
# Get robust variance-covariance matrix
robust_vcov2 <- sandwich::vcovHC(fit.pois2, type = "HC3")
tidy_model2 <- broom::tidy(fit.pois2, conf.int = TRUE, conf.level = 0.95)

# Replace standard errors with robust ones
robust_results2 <- coeftest(fit.pois2, vcov = robust_vcov2)
tidy_model2$std.error <- robust_results2[, "Std. Error"]
# Calculate RR (95% CI) with robust standard errors
tidy_poisson2 <- tidy_model2 %>%
  mutate(
    RR = exp(estimate),
    RR_low = exp(estimate - 1.96 * std.error),
    RR_high = exp(estimate + 1.96 * std.error)
  )
tidy_poisson2
```

### Model 1.3: Fully adjusted model
```{r}
form.pois3 <- as.formula(
  paste(
    "death_in_year ~ niacin_q +",
    paste(cov_full, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois3 <- glm(form.pois3,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))
summary(fit.pois3)
```

```{r}
## Poisson Model 1.3 Robust SEs
# Get robust variance-covariance matrix
robust_vcov3 <- sandwich::vcovHC(fit.pois3, type = "HC3")
tidy_model3 <- broom::tidy(fit.pois3, conf.int = TRUE, conf.level = 0.95)

# Replace standard errors with robust ones
robust_results3 <- coeftest(fit.pois3, vcov = robust_vcov3)
tidy_model3$std.error <- robust_results3[, "Std. Error"]
# Calculate RR (95% CI) with robust standard errors
tidy_poisson3 <- tidy_model3 %>%
  mutate(
    RR = exp(estimate),
    RR_low = exp(estimate - 1.96 * std.error),
    RR_high = exp(estimate + 1.96 * std.error)
  )
tidy_poisson3
```

### Model 1.4: Fully adjusted model + EMM by Sex

```{r}
form.pois4 <- as.formula(
  paste(
    "death_in_year ~ niacin_q + niacin_q*SEX +",
    paste(cov_full, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois4 <- glm(form.pois4,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))
summary(fit.pois4)
```

```{r}
## Poisson Model 1.4 Robust SEs
# Get robust variance-covariance matrix
robust_vcov4 <- sandwich::vcovHC(fit.pois4, type = "HC3")
tidy_model4 <- broom::tidy(fit.pois4, conf.int = TRUE, conf.level = 0.95)

# Replace standard errors with robust ones
robust_results4 <- coeftest(fit.pois4, vcov = robust_vcov4)
tidy_model4$std.error <- robust_results4[, "Std. Error"]
# Calculate RR (95% CI) with robust standard errors
tidy_poisson4 <- tidy_model4 %>%
  mutate(
    RR = exp(estimate),
    RR_low = exp(estimate - 1.96 * std.error),
    RR_high = exp(estimate + 1.96 * std.error)
  )
tidy_poisson4
```

```{r}
poisson_emm_results <- data.frame(
  coefs = robust_results4[c("niacin_q2", "niacin_q3", "niacin_q4", "SEX", "niacin_q2:SEX", "niacin_q3:SEX", "niacin_q4:SEX"), "Estimate"],
  se = robust_results4[c("niacin_q2", "niacin_q3", "niacin_q4", "SEX", "niacin_q2:SEX", "niacin_q3:SEX", "niacin_q4:SEX"), "Std. Error"]
)
poisson_emm_results$variance <- (poisson_emm_results$se)^2 
v <- robust_vcov4
```

## Table 3: HRs for FEMALES (SEX == 1)
```{r}
female.hr <- data.frame(
  niacin_q = c("niacin_q2", "niacin_q3", "niacin_q4"),
  variance = c(
    v["niacin_q2", "niacin_q2"] + v["niacin_q2:SEX", "niacin_q2:SEX"] + 2*v["niacin_q2", "niacin_q2:SEX"],
    v["niacin_q3", "niacin_q3"] + v["niacin_q3:SEX", "niacin_q3:SEX"] + 2*v["niacin_q3", "niacin_q3:SEX"],
    v["niacin_q4", "niacin_q4"] + v["niacin_q4:SEX", "niacin_q4:SEX"] + 2*v["niacin_q4", "niacin_q4:SEX"]),
  coef = c(
    robust_results4[c("niacin_q2"), "Estimate"] + robust_results4[c("niacin_q2:SEX"), "Estimate"],
    robust_results4[c("niacin_q3"), "Estimate"] + robust_results4[c("niacin_q3:SEX"), "Estimate"],
    robust_results4[c("niacin_q4"), "Estimate"] + robust_results4[c("niacin_q4:SEX"), "Estimate"]),
    sex = c("female", "female", "female")
)

female.hr <- female.hr |>
  mutate(
    hr = exp(coef),
    lcl = exp(coef - (qnorm(0.975)*sqrt(variance))),
    ucl = exp(coef + (qnorm(0.975)*sqrt(variance)))
  )
```

## Table 3: HRs for MALEs (SEX == 0)
```{r}
male.hr <- data.frame(
  niacin_q = c("niacin_q2", "niacin_q3", "niacin_q4"),
  variance = c(
    v["niacin_q2", "niacin_q2"],
    v["niacin_q3", "niacin_q3"],
    v["niacin_q4", "niacin_q4"]),
  coef = c(
    robust_results4[c("niacin_q2"), "Estimate"],
    robust_results4[c("niacin_q3"), "Estimate"],
    robust_results4[c("niacin_q4"), "Estimate"]),
  sex = c("male", "male", "male")
)

male.hr <- male.hr |>
  mutate(
    hr = exp(coef),
    lcl = exp(coef - (qnorm(0.975)*sqrt(variance))),
    ucl = exp(coef + (qnorm(0.975)*sqrt(variance)))
  )
```

### Combining sex-specific HRs into table
```{r}
hr.emm.sex <- rbind(female.hr, male.hr)
```

### Stratified Analysis

#### Stratifying data
```{r}
nhanes_final_poisson_male <- nhanes_final_poisson |>
  filter(SEX == 0)

table(nhanes_final_poisson_male$niacin_q)

nhanes_final_poisson_female <- nhanes_final_poisson |>
  filter(SEX == 1)
table(nhanes_final_poisson_female$niacin_q)
```

#### Full covariate set for stratified models
```{r}
cov_full_stratified <- c(
  "RACE",
  "SCHOOL",
  "MARRY",
  "URBAN",
  "BOOZE",
  "AVGSMK",
  "BMI",
  "ANGINA",
  "ARTHRITI",
  "ASPIRIN",
  "HTN_REP",
  "STROKE"
)
```

### Model 1.5: Fully-adjusted model stratified by sex - male

```{r}
form.pois5 <- as.formula(
  paste(
    "death_in_year ~ niacin_q + ",
    paste(cov_full_stratified, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois5 <- glm(form.pois5,
                 data=nhanes_final_poisson_male, 
                 family=poisson(link='log'))
summary(fit.pois5)
```

```{r}
## Poisson Model 1.5 Robust SEs
# Get robust variance-covariance matrix
robust_vcov5 <- sandwich::vcovHC(fit.pois5, type = "HC3")
tidy_model5 <- broom::tidy(fit.pois5, conf.int = TRUE, conf.level = 0.95)

# Replace standard errors with robust ones
robust_results5 <- coeftest(fit.pois5, vcov = robust_vcov5)
tidy_model5$std.error <- robust_results5[, "Std. Error"]
# Calculate RR (95% CI) with robust standard errors
tidy_poisson5 <- tidy_model5 %>%
  mutate(
    RR = exp(estimate),
    RR_low = exp(estimate - 1.96 * std.error),
    RR_high = exp(estimate + 1.96 * std.error)
  )
tidy_poisson5
```

### Model 1.6: Fully-adjusted model stratified by sex - female

```{r}
form.pois6 <- as.formula(
  paste(
    "death_in_year ~ niacin_q + ",
    paste(cov_full_stratified, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois6 <- glm(form.pois6,
                 data=nhanes_final_poisson_female, 
                 family=poisson(link='log'))
summary(fit.pois6)
```

```{r}
## Poisson Model 1.6 Robust SEs
# Get robust variance-covariance matrix
robust_vcov6 <- sandwich::vcovHC(fit.pois6, type = "HC3")
tidy_model6 <- broom::tidy(fit.pois6, conf.int = TRUE, conf.level = 0.95)

# Replace standard errors with robust ones
robust_results6 <- coeftest(fit.pois6, vcov = robust_vcov6)
tidy_model6$std.error <- robust_results6[, "Std. Error"]
# Calculate RR (95% CI) with robust standard errors
tidy_poisson6 <- tidy_model6 %>%
  mutate(
    RR = exp(estimate),
    RR_low = exp(estimate - 1.96 * std.error),
    RR_high = exp(estimate + 1.96 * std.error)
  )
tidy_poisson6
```

#### Exporting results
```{r}
# deaths per NQ
nhanes_deaths_niacin_q <- table(nhanes_final_poisson$death_in_year, nhanes_final_poisson$niacin_q)
nhanes_deaths_niacin_q

# Export CSVs
write.csv(tidy_poisson1, "nhanes_poisson_1_results.csv", row.names = FALSE)
write.csv(tidy_poisson2, "nhanes_poisson_2_results.csv", row.names = FALSE)
write.csv(tidy_poisson3, "nhanes_poisson_3_results.csv", row.names = FALSE)
write.csv(hr.emm.sex, "nhanes_poisson_4_results.csv", row.names = FALSE)
write.csv(tidy_poisson4, "nhanes_poisson_4_1_results.csv", row.names = FALSE)
write.csv(tidy_poisson5, "nhanes_poisson_5_results.csv", row.names = FALSE)
write.csv(tidy_poisson6, "nhanes_poisson_6_results.csv", row.names = FALSE)
```

### Non-linearity in baseline hazard:

In this check for non-linearity, we already decided to use a restricted spline model. 

Model 1.3.1: no term for age at follow-up
Model 1.3.2: linear term for age at follow up 
Model 1.3.3: spline term for age at follow up (model we have used for primary analysis)

* Note: Model 1.3.3 = Model 1.3 

#### Specifying models
```{r}
## Model 1.3.1
form.pois1.3.1 <- as.formula(
  paste(
    "death_in_year ~ niacin_q +",
    paste(cov_full, collapse = " + "),
    "+ offset(log_pt)"
  )
)

fit.pois1.3.1 <- glm(form.pois1.3.1,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))

## Model 1.3.2
form.pois1.3.2 <- as.formula(
  paste(
    "death_in_year ~ niacin_q +",
    paste(cov_full, collapse = " + "),
    "+ age_in_year + offset(log_pt)"
  )
)

fit.pois1.3.2 <- glm(form.pois1.3.2,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))

## Model 1.3.3 - restricted, cubic spline for age_in_year
form.pois1.3.3 <- as.formula(
  paste(
    "death_in_year ~ niacin_q +",
    paste(cov_full, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois1.3.3 <- glm(form.pois1.3.3,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))
```

#### LRTests
```{r}
# LRT #1: Compare 1.3.1 [null] to 1.3.2 [alternative]
lrtest(fit.pois1.3.1, fit.pois1.3.2)

# LRT #2: Compare 1.3.1 [null] to 1.3.3 [alternative]
lrtest(fit.pois1.3.1, fit.pois1.3.3)

# LRT #3: Compare 1.3.2 [null] to 1.3.3 [alternative]
lrtest(fit.pois1.3.2, fit.pois1.3.3)
```

### Non-linearity in NIACIN:

In this check for non-linearity, we already decided to use a restricted spline model. 

Model 1.3.4: linear term for niacin
Model 1.3.5: spline term for age at follow up (model we have used for primary analysis)

#### Specifying models
```{r}
## Model 1.3.4
form.pois1.3.4 <- as.formula(
  paste(
    "death_in_year ~ NIACIN +",
    paste(cov_full, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois1.3.4 <- glm(form.pois1.3.4,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))

## Model 1.3.5
form.pois1.3.5 <- as.formula(
  paste(
    "death_in_year ~ ns(NIACIN, df = 4) +",
    paste(cov_full, collapse = " + "),
    "+ ns(age_in_year, df = 4) + offset(log_pt)"
  )
)

fit.pois1.3.5 <- glm(form.pois1.3.5,
                 data=nhanes_final_poisson, 
                 family=poisson(link='log'))
```

#### LRTests
```{r}
# LRT #4: Compare 1.3.4 [null] to 1.3.5 [alternative]
lrtest(fit.pois1.3.4, fit.pois1.3.5)
```

## Sensitivity Analysis: Cox PH Model

### Age-adjusted
```{r}
# Age adjusted model
fit.cox.unadj <- coxph(
  Surv(age_tstart, age_tend, event) ~ niacin_q,
  data = nhanes_model,
  ties = 'efron'
)
summary(fit.cox.unadj)

cox.age <- tidy(
  fit.cox.unadj,
  exponentiate = TRUE,
  conf.int = TRUE
) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    HR = estimate,
    CI_lower = conf.low,
    CI_upper = conf.high,
    p_value = p.value
  )
```

### Minimally adjusted 
```{r}
# Minimally adjusted model
fit.cox.min.adj <- coxph(
  as.formula(
  paste(
    "Surv(age_tstart, age_tend, event) ~ niacin_q +",
    paste(cov_basic, collapse = " + ")
  )),
  data = nhanes_model,
  ties = 'efron'
)
summary(fit.cox.min.adj)

cox.minimal <- tidy(
  fit.cox.min.adj,
  exponentiate = TRUE,
  conf.int = TRUE
) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    HR = estimate,
    CI_lower = conf.low,
    CI_upper = conf.high,
    p_value = p.value
  )
```

### Fully-adjusted 
```{r}
# Fully adjusted model
fit.cox.full.adj <- coxph(
  as.formula(
  paste(
    "Surv(age_tstart, age_tend, event) ~ niacin_q +",
    paste(cov_full, collapse = " + ")
  )),
  data = nhanes_model,
  ties = 'efron'
)
summary(fit.cox.full.adj)

cox.full <- tidy(
  fit.cox.full.adj,
  exponentiate = TRUE,
  conf.int = TRUE
) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    HR = estimate,
    CI_lower = conf.low,
    CI_upper = conf.high,
    p_value = p.value
  )
```

### Exporting Cox PH Results
```{r}
write_csv(cox.age, "cox_ageadj_results.csv")
write_csv(cox.minimal, "cox_minimal_results.csv")
write_csv(cox.full, "cox_full_results.csv")
```

### Schoenfeld Residuals Check

#### For Niacin Quartile
```{r}
#checking Schoenfeld Residuals
check.pha<-cox.zph(fit.cox.full.adj)
ggcoxzph(check.pha[1]) #AB: PH holds - checked with Mingyang
```

#### For all covariates
```{r}
form.test.cox <- as.formula(
  paste(
    "Surv(age_tstart, age_tend, event) ~ niacin_q +",
    paste(cov_full, collapse = " + ")
  )
)

test.cox <- coxph(
  form.test.cox,
  data = nhanes_model,
  ties = "efron"
)

cox.test <- cox.zph(test.cox)
plot_names <- rownames(cox.test$table)
plot_names <- plot_names[plot_names != "GLOBAL"]

pdf("schoenfeld_plots.pdf", width = 10, height = 16)

par(mfrow = c(7, 2), mar = c(3, 3, 2, 1))

for (i in seq_along(plot_names)) {
  plot(cox.test, var = i, main = plot_names[i])
}

dev.off()

par(mfrow = c(1, 1))
```

## Table 1 Code 

Purpose: Describing relevant basic characteristics of study participants according to exposure categories of your choice

```{r}
table.1<-nhanes_final_2 %>% 
  arrange(niacin_q) %>% 
  mutate(niacin_cat=factor(niacin_q,levels=c(1,2,3,4)),
        sex=factor(SEX, labels=c("male", "female"), levels=c(1,2)), 
         age= AGEYRS,
         angina=factor(ANGINA, levels=c(0,1)),
         arthritis = factor(ARTHRITI, levels=c(0,1)),
         aspirin = factor(ASPIRIN, levels=c(0,1)),
         ever_pregnant=factor(EVPREG,levels=c(0,1)), # might need to show it only among females
         hypertension=factor(HTN_REP,levels=c(0,1)),
         marital_status=factor(MARRY, 
                               labels=c("Under 17", "Married", "Widowed",
                                        "Divorced", "Separated", "Never Married",
                                        "Blank, but applicable"), 
                               levels=c(1,2,3,4,5,6,8)),
         race=factor(RACE,
                     labels=c("White","Black","Other"),
                     levels=c(1,2,3)),
         urban=factor(URBAN,
                      labels=c("Yes", "No"), levels=c(1,2)),
         school=factor(SCHOOL,
                       labels=c("Elementary", "Some high school", 
                                      "High school", "Some college", "College"), 
                               levels=c(1,2,3,4,5)),
         alcohol=as.numeric(BOOZE),
         recex=factor(RECEX, labels=c("High","Medium","Low"),levels=c(1,2,3)),
         smoking=as.numeric(AVGSMK),
         bmi=as.numeric(BMI))

vars <- (c("sex", "race", "age","angina","arthritis", 
           "hypertension", "aspirin",
           "school",
           "marital_status", "urban",
            "alcohol", "recex", "smoking",
           "bmi")) # include more covariates as needed

# Table 1 needs complete case

t1 <- CreateTableOne(vars = vars, data = table.1, strata = as.factor("niacin_q"))
options(digits = 2)
print(t1, showAllLevels = TRUE)
```

## Appendix (for fun): forest plot

```{r, eval = FALSE}
forest_df <- bind_rows(
  tidy_poisson1 %>% mutate(model = "Model 1"),
  tidy_poisson2 %>% mutate(model = "Model 2"),
  tidy_poisson3 %>% mutate(model = "Model 3")
) %>%
  filter(term %in% c("niacin_q2", "niacin_q3", "niacin_q4")) %>%
  mutate(term = recode(term,
    "niacin_q2" = "Q2 vs Q1",
    "niacin_q3" = "Q3 vs Q1",
    "niacin_q4" = "Q4 vs Q1"
  ))


ggplot(forest_df,
       aes(x = RR,
           y = model,
           xmin = RR_low,
           xmax = RR_high,
           color = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbarh(position = position_dodge(width = 0.5), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Hazard Ratio (95% CI)",
    y = "Model",
    title = "Association Between Niacin Quartile and Mortality"
  ) +
  theme_bw()
```





